Structural Breaks in Inflation Dynamics within the European Monetary Union

### 2012-03-18 pkg/paper/Paper/EmpEc.Rnw
updated version
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/glogis@126 b679b2f7-6539-45d0-a605-5ddcb976cde9

All empirical analyses are based on seasonally adjusted inflation rates (in percent)
for \Sexpr{ncol(hicps)}~countries
(\Sexpr{gsub("UK", "United Kingdom", gsub("CzechRep", "Czech Rep", paste(colnames(hicps), collapse = ", "), fixed = TRUE), fixed = TRUE)})
in the time period from the early or mid-1990s to the end of 2010.
The original data source are seasonally unadjusted HICP series provided by the
\citet{oecd} from January~1990 (if available) to December~2010, that were transformed
to inflation rates using log-returns and subsequent seasonal adjustment
via X-12-ARIMA \citep{Findley+Monsell+Bell:1998}.

@@

Finally, Section~\ref{sec:fcrisis} focuses on the question of changes
in the inflation dynamics during the 2008/2009 financial crisis.

@@

As a first step towards the investigation of inflation convergence, the
structural breaks in the inflation dynamics within each country are
investigated using the methods introduced in Section~\ref{sec:model}.
Table~\ref{tab:results1} reports the estimated breakpoints along with the
segment-specific moments (mean, variance, skewness) implied by the corresponding
parameter estimates. Furthermore, entry dates for ERM/ERM~II and the euro are
included as additional information. For ease of interpretation, countries with
similar patterns of change are grouped together and ordered alphabetically
within each group. If there is only a single segment, the sup$\mathit{LM}$ test
was non-significant at the 5\% level. In significant cases, the number of
breakpoints was selected via the LWZ criterion and subsequently all parameters
(breakpoints and segment-specific GL parameters) were estimated by ML. No
HAC~correction was employed as there was only negligible autocorrelation in the
empirical scores after inclusion of the breakpoints (if any).
In Table~\ref{tab:results1} the countries are grouped according to either a
similar timing of the changes in inflation dynamics or according to economic similarities, 
as in the case of the former communist countries in Eastern Europe.

The first group that draws particular attention consists of the former
DEM (Deutsche Mark) zone countries, with the exception of Austria and
Luxembourg. All these countries had and still have a very low inflation rate,
both in terms of mean and variance, and their inflation dynamics
did not change significantly. This is plausible from an economic point of view
since the European central bank (ECB) is modeled after the Deutsche Bundesbank and the ECB did not
-- until recently -- diverge much from the monetary policy of the former. Thus,
the EMU did have no effect on inflation dynamics since the policies in
these countries were already much in line with Germany and hence the ECB.

@@

Austria, Finland, and France adjusted to the requirements of the EMU during the
first phase of convergence. In these countries,
the result of the adjustment process was a lower inflation rate and an overall
decrease in the variance.

The next group consists of three of the PIIGS countries and Luxembourg. Greece and
Portugal adjusted to the ERM shortly before or after its introduction.
Italy was successful in reducing its inflation rate, whereas Luxembourg
experienced a strong increase in its formerly very low inflation rate, both
in terms of mean and variance, also associated with a shift in skewness towards the
left. Apart from Luxembourg, the countries that had a structural break during the second 
phase of convergence were also successful in reducing mean inflation rates.
After the changes in this group, all countries have rather similar mean inflation
rates, which are closer to the levels of the previous two groups.

The fourth subgroup consists of two of the PIIGS countries, namely Ireland and Spain.  
There is a break in both countries in the summer of 2008 that captures the effects of
the financial crisis and the severe problems concerning the real estate situation.
In both countries, this had a strong deflationary impact, with deflation in Ireland
over an extended period of time. In Spain, which was affected by a housing bubble of  
roughly the same size as Ireland, the housing prices did not adjust so abruptly and  
so strongly.

The next group consists of two countries that, after a very short participation in the
ERM, decided not to be part of it and hence also not to participate in the EMU:
the United Kingdom and Sweden. Both show clear breaks in the early 1990s which can be
traced back to economic crises: the currency crisis of the United Kingdom cumulating
into the ``Black Wednesday'' in 1992 and the banking crisis in Sweden during the early
1990s.

The last group consists of the Eastern European countries: Czech Republic,
Estonia, Hungary, Poland, Slovakia, and Slovenia. In almost all of these
countries, the mean and the variance of their inflation rates declined in the
late 1990s or early 2000s. Czech Republic, Estonia, and Hungary experienced a break in 1997/1998
resulting from efforts to curb in inflation by decreasing the growth of the money 
supply. The timing of the break thus indicates the date when the the biggest
transitionally shocks on the way towards free market economies have been overcome. In Poland, this
transition took two more years. Slovakia and Slovenia are somewhat different: On the one hand,
they started with lower mean inflation rates and on the other hand, their inflation dynamics
both changed roughly one year prior to their entry into the ERM~II.

>>>>

All empirical analyses are based on seasonally adjusted monthly inflation rates (in percent)
for \Sexpr{ncol(hicps)}~countries
(\Sexpr{gsub("UK", "United Kingdom", gsub("CzechRep", "Czech Rep", paste(colnames(hicps), collapse = ", "), fixed = TRUE), fixed = TRUE)})
in the time period from the early or mid-1990s to the end of 2010.
The original data source are seasonally unadjusted HICP series provided by the
\citet{oecd} from January~1990 (if available) to December~2010, that were transformed
to inflation rates using log-returns and subsequent seasonal adjustment
via X-12-ARIMA \citep{Findley+Monsell+Bell:1998}.

@@

Section~\ref{sec:fcrisis} focuses on the question of changes
in the inflation dynamics during the 2008/2009 financial crisis. Finally, the model fit and related issues will be discussed in Section~\ref{sec:fit}

@@

As a first step towards the investigation of inflation convergence, the
structural breaks in the inflation dynamics within each country are
investigated using the methods introduced in Section~\ref{sec:model}.
Table~\ref{tab:results1} reports the estimated breakpoints along with the
segment-specific moments (mean, variance, skewness) implied by the corresponding
parameter estimates. Furthermore, entry dates for ERM/ERM~II and the euro are
included as additional information. For ease of interpretation, countries with
similar patterns of change are grouped together and ordered alphabetically
within each group. If there is only a single segment, the sup$\mathit{LM}$ test
was non-significant at the 5\% level. In significant cases, the number of
breakpoints was selected via the LWZ criterion and subsequently all parameters
(breakpoints and segment-specific GL parameters) were estimated by ML. 
The $p$~values from the sup$\mathit{LM}$ test with and without HAC correction are reported in Table~\ref{tab:diag} 
along with further diagnostics for each of the fitted segments concerning goodness of fit and autocorrelation. 
As autocorrelation was very low in most of the segments (typically below 0.4 and non-significant, with a few exceptions), 
we focus on the results from the analysis without HAC correction but point out if there are noteworthy deviations. 
In both Table~\ref{tab:results1} and~\ref{tab:diag} the countries are grouped  according to either a
similar timing of the changes in inflation dynamics or according to economic similarities, 
as in the case of the former communist countries in Eastern Europe.

The first group that draws particular attention consists of the former
DEM (Deutsche Mark) zone countries, with the exception of Austria and
Luxembourg. All these countries had and still have a very low inflation rate,
both in terms of mean and variance, and their inflation dynamics
did not change significantly. This is plausible from an economic point of view
since the European central bank (ECB) is modeled after the Deutsche Bundesbank and the ECB did not
-- until recently -- diverge much from the monetary policy of the former. Thus,
the EMU did have no effect on inflation dynamics since the policies in
these countries were already much in line with Germany and hence the ECB. Table~\ref{tab:diag} does not 
suggest any problems in this segment.

@@

Austria, Finland, and France adjusted to the requirements of the EMU during the
first phase of convergence. However, with HAC correction, the very short segment of France would not be
segregated, thereby suggesting that France belongs to the group of countries which did not change at all.
In these countries,
the result of the adjustment process was a lower inflation rate and an overall
decrease in the variance.

The next group consists of three of the PIIGS countries and Luxembourg. Greece and
Portugal adjusted to the ERM shortly before or after its introduction.
Italy was successful in reducing its inflation rate, whereas Luxembourg
experienced a strong increase in its formerly very low inflation rate, both
in terms of mean and variance, also associated with a shift in skewness towards the
left. Apart from Luxembourg, the countries that had a structural break during the second 
phase of convergence were also successful in reducing mean inflation rates.
After the changes in this group, all countries have rather similar mean inflation
rates, which are closer to the levels of the previous two groups. The diagnostics for model fits in 
this group are unproblematic.

The fourth subgroup consists of two of the PIIGS countries, namely Ireland and Spain.  
There is a break in both countries in the summer of 2008 that captures the effects of
the financial crisis and the severe problems concerning the real estate situation.
In both countries, this had a strong deflationary impact, with deflation in Ireland
over an extended period of time. For the middle segment of Spain, the Ljung-box shows that not all
of the autocorrelation has been captured. In Spain, which was affected by a housing bubble of  
roughly the same size as Ireland, the housing prices did not adjust so abruptly and  
so strongly.

The next group consists of two countries that, after a very short participation in the
ERM, decided not to be part of it and hence also not to participate in the EMU:
the United Kingdom and Sweden. Both show clear breaks in the early 1990s which can be
traced back to economic crises: the currency crisis of the United Kingdom cumulating
into the ``Black Wednesday'' in 1992 and the banking crisis in Sweden during the early
1990s. With regard to the last segment for both countries, the Ljung-box suggests that the data 
in this segment is independently distributed. Anyhow, the HAC correction would also suggest two breaks 
for both series.

The last group consists of the Eastern European countries: Czech Republic,
Estonia, Hungary, Poland, Slovakia, and Slovenia. In almost all of these
countries, the mean and the variance of their inflation rates declined in the
late 1990s or early 2000s. For Estonia and Slovakia, the $\chi^2$ goodness-of-fit suggests that the generalized logistic distribution
might not be the most suitable distribution for this purpose. Also, compared to other country groups, the 
Eastern European group shows the highest occurrence of non-captured autocorrelation. Nevertheless, compared to other 
inflation rates, these series show the most dynamic behaviour.

Czech Republic, Estonia, and Hungary experienced a break in 1997/1998
resulting from efforts to curb in inflation by decreasing the growth of the money 
supply. The timing of the break thus indicates the date when the the biggest
transitionally shocks on the way towards free market economies have been overcome. In Poland, this
transition took two more years. Slovakia and Slovenia are somewhat different: On the one hand,
they started with lower mean inflation rates and on the other hand, their inflation dynamics
both changed roughly one year prior to their entry into the ERM~II.
